Approximation Algorithms: Greedy Algorithm and Local Search

### 2024-08-28 paper/paper.tex
updated the paper

\title{Approximation Algorithms: Greedy Algorithm and Local Search}
\author{Zixuan Fan\thanks{Department of Computer Science, Technical University of Munich. {\tt ge43yeb@mytum.de}}}
\maketitle
\section{Introduction}
The theory of computer science has grown significantly throughout the past century. Among many topics, \NP-hardness has drawn the attention of many researchers. 
While there is no proof whether $\PTIME=\NP$, people tend to be interested in compensation: solving NP-hard problems with a sub-optimal solution in polynomial time, which is the motivation of approximation algorithms. This paper introduces two fundamental techniques in designing approximation algorithms: greedy algorithm and local search.
We start with some mathematical backgrounds, followed by a brief introduction of greedy and local search. 
Then, we present a few examples that exploit these techniques.

\section{Mathematical Backgrounds}
\subsection{Approximation Ratio}
Approximation algorithms attempt to solve optimization problems with a sub-optimal solution in polynomial time. 
Hence, in this context, we refer to an \NP-hard problem as an \NP-hard optimization problem instead of a decision problem. 
With the goal problems clarified, we introduce the concept of approximation ratio.
\begin{definition}[Approximation Ratio]
    Given an optimization problem, its optimal solution $Opt^*$, and a sub-optimal solution $Opt$ given by algorithm $A$, suppose the size of solutions 
    are measured by $|\cdot|$. The approximation ratio $\alpha$ of algorithm $A$ is defined as 
    \begin{align*}
        \alpha = \frac{|Opt|}{|Opt^*|}
    \end{align*}
\end{definition}
The approximation ratio $\alpha$ shows how close the solution $Opt$ is to the optimal solution $Opt^*$. For an maximization problem, $\alpha < 1$, while $\alpha > 1$
for a minimization problem. The better the approximation, the closer $\alpha$ is to 1.

\subsection{Greedy Algorithms and Local Search}
Both greedy algorithms and local search construct the solution step by step. 
In each step, the strategy makes certain decisions to optimize the result locally. 
The difference between the two techniques lies in the strategy of making decisions. 
The greedy algorithms try to make the best decision at each step, and the solution is not guaranteed to 
be \textit{feasible} at the beginning. On the other hand, local search starts with an arbitrary feasible solution
and maintains the feasibility while improving the solution. 
Hence, greedy algorithms are called \textit{primal infeasible} algorithms, while local search algorithms 
are referred to as \textit{primal feasible} algorithms.

\section{Scheduling on A Single Machine}
Scheduling, or Job Sequencing, is one of the earliest discovered NP-complete problems. 
It was amongst the first 21 NP-complete problems published by Karp in 1972. In this section, 
we examine a simplified version of the problem: only one machine is available for scheduling.
\begin{definition}[Scheduling on Single Machine]
    Given $n$ jobs to be processed with processing time $p_j$, release time $r_j \geq 0$, and deadline $d_j$ with $j = 1, \cdots, n$.
    Suppose $j$-th job is completed at time $C_j$. We define the lateness as $L_j = C_j - d_j$. The goal is to minimize the maximal lateness 
    \begin{align*}
        L_{max} = \max_{j = 1, \cdots, n} L_j
    \end{align*}
\end{definition} 
Unfortunately, the problem is \NP-hard, and it remains \NP-hard even if we apply 
a few constraints to it. \textbf{cite the source} One exceptional case is when all deadlines are non-negative: the lateness is always positive. 
We are able to give a 2-approximation algorithm for this case. The algorithm simply exploits the earliest due date rule(\textbf{EDD}). 
Just as the name suggests, whenever the machine finishes a job, it picks the job with the earliest due date from all \textbf{available} jobs.
Here, a job is \textit{available} at time $t$ if $r_j \leq t$.

To analyze the algorithm, we have to introduce a few notations. Let $S$ denote a set of jobs, $r(S) = \min_{j \in S} r_j$, $p(S) = \sum_{j \in S} p_j$, 
and $d(S) = \max_{j \in S} d_j$. While $L_{max}$ denotes the maximal lateness computed by the algorithm, $L_{max}^*$ denotes the optimal solution. 
As preparation, an interesting lower bound on the optimal solution is observed. 
\begin{lemma}
    \label{lemma:lowerbound}
    For any set of jobs $S$, $L_{max}^* \geq r(S) + p(S) - d(S)$
\end{lemma}
\begin{proof}
    Suppose job $j$ is the last finished job in the optimal schedule. The scheduling cannot start until $r(S)$. In an optimal schedule, there is no gap between the processing of jobs, and all jobs 
    are processed consecutively in $p(S)$. Hence, $j$ cannot be finished until $r(S) + p(S)$. In addition, we have $d_j \leq d(S)$ by definition. It follows 
    \begin{align*}
        L_{max}^* \geq L_j = C_j - d_j \geq r(S) + p(S) - d_j \geq r(S) + p(S) - d(S)
    \end{align*} 
\end{proof}
We then take a closer look at the \textbf{EDD} rule. Suppose the release time $r_j$, processing time $p_j$, and the deadline $d_j$ are encoded as a ternary tuple $(r_j, p_j, d_j)$.
The only part that requires heavy computation for \textbf{EDD} rule is choosing the job with the earliest due date.
The easiest implementation is to preserve a linked list to store the information of each job, which takes $\bigo{n^2}$ time in total. 
A more efficient implementation takes advantage of a FIFO queue, which takes only $\bigo{n}$ time. In either case, 
the algorithm runs in polynomial time. This directly leads to the following theorem.
\begin{theorem}
    The EDD-rule algorithm runs in polynomial time. 
\end{theorem}
What we care more about in the context of approximation algorithms is the approximation ratio. We then show that the algorithm yields the promised approximation ratio.
\begin{theorem}
    The EDD-rule algorithm yields a 2-approximation ratio.
\end{theorem}
\begin{proof}
    Suppose $j$ is the job with the maximal lateness in the schedule generated by the \textbf{EDD} rule.
    Let $t$ be the earliest time for the machine to keep processing jobs continuously in the interval $[t, C_j]$.
    Let $S$ be the set of jobs that are processed in this interval. In our assumption, $r(S) = t$ is guaranteed. 
    Suppose for contradiction that $r(S) > t$, the processing would not be able to start at time $t$. 
    Similarly, if $r(S) < t$, the processing would start earlier than $t$. Both cases contradict the assumption.
    Furthermore, $p(S) = C_j - t$, for the processing is continuous, and it stops at time $C_j$. Applying lemma \cref{lemma:lowerbound},
    we obtain 
    \begin{align*}
        L_{max}^* \geq r(S) + p(S) - d(S) = t + C_j - t - d(S) = C_j - d(S) \geq C_j
    \end{align*}
    Additionally, it holds $d_j < d(S)$ by definition. Applying the same lemma, we have
    \begin{align*}
        L_{max}^* \geq r(S) + p(S) - d(S) \geq -d(S) \geq -d_j
    \end{align*}
    Observe that $L_{max} = C_j - d_j$. Hence, it holds for this schedule
    \begin{align*}
        L_{max} = C_j - d_j \leq 2L_{max}^*
    \end{align*}
    which justifies the 2-approximation ratio.
\end{proof}

>>>>

\title{Approximation Algorithms: Greedy Algorithm and Local Search}
\author{Zixuan Fan\thanks{Department of Computer Science, Technical University of Munich. {\tt ge43yeb@mytum.de}}}
\maketitle
\section{Introduction}
The theory of computer science has grown significantly throughout the past century. Among many topics, \NP-hardness has drawn the attention of many researchers. 
While there is no proof whether $\PTIME=\NP$, people tend to be interested in alternatives: solving NP-hard problems with a sub-optimal solution in polynomial time, 
the approximation algorithms. This paper introduces two fundamental techniques in designing approximation algorithms: greedy algorithm and local search algorithms.
We start with some mathematical backgrounds, followed by a brief introduction of two techniques. 
Finally, we present a few examples that exploit these techniques.

\section{Mathematical Backgrounds}
\subsection{Approximation Ratio}
Approximation algorithms attempt to solve optimization problems with a sub-optimal solution in polynomial time. 
Hence, in this context, we refer to an \NP-hard problem as an \NP-hard optimization problem instead of a decision problem. 
To evaluate the quality of approximation, we introduce approximation ratio.
\begin{definition}[Approximation Ratio]
    Let $Opt^*$ be the optimal solution of an optimization problem, and let $Opt$ be its solution given by algorithm $A$. 
    Suppose the size of solutions is measured by $|\cdot|$. The approximation ratio $\alpha$ of algorithm $A$ is defined as 
    \begin{align*}
        \alpha = \frac{|Opt|}{|Opt^*|}
    \end{align*}
\end{definition}
The approximation ratio $\alpha$ shows how close the solution $Opt$ is to the optimal solution $Opt^*$. For an maximization problem, $\alpha < 1$, while $\alpha > 1$
for a minimization problem.

\subsection{Greedy Algorithms and Local Search}
Both greedy and local search algorithms construct the solution step by step. 
In each step, the strategy makes certain decisions to optimize the result locally. 
Their difference is where the algorithm starts. 
The greedy algorithm usually starts with an empty solution, which may not be feasible. 
On the other hand, local search starts with an arbitrary feasible solution
and maintains the feasibility while improving the solution. 
Hence, greedy algorithms are called \textit{primal infeasible} algorithms, while local search algorithms 
are referred to as \textit{primal feasible} algorithms.

\section{Scheduling on A Single Machine}
Scheduling, or Job Sequencing, is one of the earliest discovered \textbf{NP}-complete problems. 
It was amongst the first 21 \textbf{NP}-complete problems published by Karp in 1972. In this section, 
we examine a simplified version of the problem: only one machine is available for processing.
\begin{definition}[Scheduling on Single Machine]
    Given $n$ jobs to be processed with processing time $p_j$, release time $r_j \geq 0$, and deadline $d_j$ with $j = 1, \cdots, n$.
    Suppose $j$-th job is completed at time $C_j$. We define the lateness as $L_j = C_j - d_j$. The goal is to minimize the maximal lateness 
    \begin{align*}
        L_{max} = \max_{j = 1, \cdots, n} L_j
    \end{align*}
\end{definition} 
Unfortunately, the problem is \NP-hard, and it remains \NP-hard even if we apply 
a few constraints to it. One exceptional case is when all deadlines are non-negative: the lateness is always positive. 
We are able to give a 2-approximation algorithm for this case. The algorithm simply exploits the earliest due date rule know as the \textbf{EDD} rule. 
Just as the name suggests, whenever the machines is idle, it picks the job with the earliest due date from all \textit{available} jobs.
Here, a job is \textit{available} at time $t$ if $r_j \leq t$.

To analyze the algorithm, we have to introduce a few notations. Let $S$ denote a subset of all jobs, $r(S) = \min_{j \in S} r_j$, $p(S) = \sum_{j \in S} p_j$, 
and $d(S) = \max_{j \in S} d_j$. While $L_{max}$ denotes the maximal lateness computed by the algorithm, $L_{max}^*$ denotes the optimal solution. 
An interesting lower bound on the optimal solution can be observed. 
\begin{lemma}
    \label{lemma:lowerbound}
    For any subset of jobs $S$, $L_{max}^* \geq r(S) + p(S) - d(S)$
\end{lemma}
\begin{proof}
    Suppose job $j$ is the last finished job in the optimal schedule. The scheduling cannot start until $r(S)$. In an optimal schedule, there is no gap between the processing of jobs, and all jobs 
    are processed consecutively in $p(S)$. Hence, $j$ cannot be finished until $r(S) + p(S)$. In addition, we have $d_j \leq d(S)$ by definition. It follows 
    \begin{align*}
        L_{max}^* \geq L_j = C_j - d_j \geq r(S) + p(S) - d_j \geq r(S) + p(S) - d(S)
    \end{align*} 
\end{proof}
We then take a closer look at the \textbf{EDD} rule. Suppose the release time $r_j$, processing time $p_j$, and the deadline $d_j$ are encoded as a ternary tuple $(r_j, p_j, d_j)$.
The only part that requires heavy computation for \textbf{EDD} rule is choosing the job with the earliest due date.
The easiest implementation is to preserve a linked list to store the information of each job, which takes $\bigo{n^2}$ time in total. 
A more efficient implementation takes advantage of a FIFO queue, which takes only $\bigo{n}$ time. In either case, 
the algorithm runs in polynomial time. This directly leads to the following theorem.
\begin{theorem}
    The EDD-rule algorithm runs in polynomial time. 
\end{theorem}
What we care more about in the context of approximation algorithms is the approximation ratio. We then show that the algorithm yields the promised approximation ratio.
\begin{theorem}
    The EDD-rule algorithm yields a 2-approximation ratio.
\end{theorem}
\begin{proof}
    Suppose $j$ is the job with the maximal lateness in the schedule generated by the \textbf{EDD} rule.
    Let $t$ be the earliest time for the machine to keep processing jobs continuously in the interval $[t, C_j]$.
    Let $S$ be the set of jobs that are processed in this interval. In our assumption, $r(S) = t$ is guaranteed. 
    Suppose for contradiction that $r(S) > t$, the processing would not be able to start at time $t$. 
    Similarly, if $r(S) < t$, the processing would start earlier than $t$. Both cases contradict the assumption.
    Furthermore, $p(S) = C_j - t$, for the processing is continuous, and it stops at time $C_j$. Applying lemma \cref{lemma:lowerbound},
    we obtain 
    \begin{align*}
        L_{max}^* \geq r(S) + p(S) - d(S) = t + C_j - t - d(S) = C_j - d(S) \geq C_j
    \end{align*}
    Additionally, it holds $d_j < d(S)$ by definition. Applying the same lemma, we have
    \begin{align*}
        L_{max}^* \geq r(S) + p(S) - d(S) \geq -d(S) \geq -d_j
    \end{align*}
    Observe that $L_{max} = C_j - d_j$. Hence, it holds for this schedule
    \begin{align*}
        L_{max} = C_j - d_j \leq 2L_{max}^*
    \end{align*}
    which justifies the 2-approximation ratio.
\end{proof}
